---
title: "Homework 1"
author: "Ian Effendi"
output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{rmarkdown}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Certication

> I certify that I indeed finished reading Ch. 2 from *An Introduction to Statistical Learning*, by James Gareth, Daniela Witten, Trevor Hastie, Robert Tibshirani
> --- Ian Effendi (`iae2784@rit.edu`)

The assignment tasks are completed below and provided in a brief report that highlights used R code and results.

```{r setup}
# library(RIT.STAT745.HW1)
devtools::load_all("./")
```

## 1. Read in the data from the object, which was saved with `save(galaxies, file="galaxies.RData")` `R` command.

```{r}
load(file = "data/galaxies.RData")
?galaxies
```

## 2. Perform an EDA on the data.

### Describe shape of data

TODO: Count number of rows.
TODO: Count number of factors.

### Describe categories

TODO: Describe each category and typical summary statistics for each.

### Summary statistics

TODO: Additional summary statistics can be provided here.

## 3. Fit a linear, no-intercept model (called Hubble's law).

$velocity = \beta_1 * distance + \epsilon$

## 4. Assess the quality of the model fit, but do not explore other models.

## 5. Estimate $\beta_1$ (called Hubble's constant), including units. Hubble's constant is given in $km * sec^{-1} * Mpc^{-1}$. A `mega-parsec` ($Mpc$) is $3.086 * 10^{19} km$. Velocity data is given in $\frac{km}{s}$ and distance in $Mpc$.

TODO: Calculate $\beta_{1}$.

## 6. Find $\beta_{1}^{-1}$ (which approximates the age of the universe) in seconds, and then transform it into `years`.

TODO: Calculate $\beta_{1}^{-1}$.
