---
# Document metadata information:
title: "Homework 3 Report"
subtitle: Classification of liver malfunction severity (LDA)
author: Ian Effendi \\ iae2784@rit.edu
date: "`r format(Sys.time(), '%B %d, %Y')`"
# Output formats:
output:
  # Output as a previewable HTML Notebook.
  html_notebook:
    theme: united
    highlight: tango
    df_print: paged
    toc: true
    toc_depth: 2
  # Output as an rmarkdown::pdf_document()
  pdf_document:
    keep_tex: true
    highlight: tango
    fig_caption: true
    df_print: kable
    toc: true
    toc_depth: 3
# pandoc -> pdfLaTeX rendering options:
fontsize: 11pt
geometry:
  - margin=1in
  - heightrounded
documentclass: scrartcl
papersize: a4
urlcolor: violet
toccolor: blue
# Custom knit function.
#knit: (function(inputFile, encoding){
#  rmarkdown::render(inputFile, encoding = encoding, 
#  output_dir = "vignettes/out", 
#  output_format="all") })
---

## Certification

> I certify that I indeed finished reading Ch. 4 from *An Introduction to Statistical Learning*, by James Gareth, Daniela Witten, Trevor Hastie, Robert Tibshirani.

## Overview

```{r setup, echo=FALSE, include=FALSE, results='hide'}

# Comments that begin with '## ----' can be used
# to separate content. R Studio provides them as 
# little bookmarks that you can use.

## ---- Hidden Variables ----
.VERBOSE   <- TRUE
.TARGET <- "Liver"
.FILEPATHS <- list(
  utils      = list(
      .            = "R/utils.R",
      dependencies = "R/utils/dependencies.R",
      paths        = "R/utils/paths.R",
      printf       = "R/utils/printf.R"
  ),
  packages   = "R/packages.R",
  build      = list(
      .            = "R/build.R",
      dataset      = "R/build/dataset.R",
      report       = "R/build/report.R"
  ),
  analysis   = list(
      .            = "R/analysis.R",
      setup        = "R/analysis/setup.R",
      eda          = "R/analysis/eda.R",
      model        = "R/analysis/model.R",
      metrics      = "R/analysis/metrics.R",
      validation   = "R/analysis/validation.R"
  )
)

## ---- Default Settings (Knit) ----
knitr::opts_knit$set( 
  root.dir = here::here()
)

## ---- Default Settings (Chunk) ----
knitr::opts_chunk$set(
  
  ## ---- Options (Execution) ----
  eval        = TRUE,   # By default, turn off chunk evaluation.
  cache       = FALSE,  # By default, cache all chunk results.
  error       = FALSE,   # Stop R when an error is raised.
  
  ## ---- Options (Figure) ----
  fig.path    = 'figure/',
  fig.keep    = 'all',
  fig.align   = 'center',
  fig.width   = 8,
  fig.height  = 8,
  dpi         = 600,
  
  ## ---- Options (Formatting) ----
  results     = 'hold',
  collapse    = FALSE,
  strip.white = FALSE,  # Remove blank lines in output.
  tidy        = FALSE,
  size        = 'tiny',
  R.options   = list(width = 60),

  ## ---- Options (Display) ----
  include     = TRUE,
  echo        = TRUE,   
  message     = FALSE,
  warning     = FALSE
)

## ---- Utility Functions ----

#' Use relative path to source an
#' external script in its entirety.
#'
#' @param path Relative path to file.
#' @param ... Additional parameters passed to the `source` call.
#' @param verbose Print messages to stderr.
import.source <- function(path, ..., verbose = .VERBOSE) {
  if (verbose) { message(sprintf("Sourcing external script: '%s'...", basename(path))) }
  source(here::here(path), ..., echo = .VERBOSE)
}
```

```{r import-packages, include=FALSE, echo=FALSE, message=FALSE}
## ---- Source (R/packages.R) ----
import.source(.FILEPATHS$packages)
```


```{r import-analysis, include=FALSE, echo=FALSE, message=FALSE}
## ---- Source (R/analysis.R) ----
import.source(.FILEPATHS$analysis$.)
```

<!-- This section provides an overview of the report -->

In this assignment we will:

- Perform exploratory data analysis (*EDA*) on the dataset.
- Fit and analyze a linear discriminant analysis (*LDA*) model on the dataset.
- Perform multiple cross-validation tasks at different $k$-fold values ($k=3,\,k=10$).

## ELT

*Much of the __extract__, __load__, and __transform__ (__ELT__) process from the previous report has been revised for this assignment. Notably, a `make.dataset()` function streamlines the process of parsing the source `Liver.txt` file into a compatible `data.frame`.*

```{r analysis-setup, echo=TRUE, message=TRUE}
# Import the dataset as variable called "Liver".
setup.analysis(target = "Liver")
```

The `setup.analysis()` function imports a dataset into the global environment with the name provided to the `target =` argument.

## EDA {.tabset}

*This section reviews the `Liver` dataset. In an improvement over the previous report, it now incorporates an assessment of feature correlations.*

### Response Encoding

```{r analysis-eda-recode}
# Recode the response.
liver <- Liver %>% make.response()
summary(liver)
```

```{r analysis-eda-truth, echo=1, results='hide'}
truth <- (liver$severity == "Severe")
summary(truth)
```

<!-- PAGE BREAK -->
\newpage

### Feature Correlations

```{r analysis-eda-summary, eval=FALSE, echo=TRUE}
liver.info <- analysis.eda(Liver, truth)
summary(liver.info)
```

```{r analysis-eda-summary-presentation, echo=FALSE, results='hold'}
liver.info <- analysis.eda(Liver, truth)
print(unlist(liver.info$shape, use.names = TRUE))
println("Label counts: ");
print(unlist(liver.info$n_labels, use.names = TRUE))
println("Class prior probabilities: ");
options(digits = 4)
print(unlist(liver.info$p_labels, use.names = TRUE))
```
```{r analysis-eda-corrplot}
# Plot using corrplot::corrplot wrapper.
liver.corrplot <- corr.plot(
  liver, sig.level = 0.05, insig = "blank",
  title = "Correlation Plot",
  mar = c(1,1,2,1)
)
```

The data behind the above correlation plot is provided below:

```{r analysis-eda-correlation, echo=FALSE, results='hold'}
println("Feature correlation matrix: ")
print(liver.corrplot$corr)
```
### Baseline Model Analysis

```{r analysis-glm-baseline, results='hide'}
# TODO: Replace glm with lda model.
liver.baseline <- analysis.clf(
  liver, truth,
  algorithm = glm,
  params = list(
    formula = severity ~ .,
    family = "binomial"
  ),
  pos = "Severe", neg = "Not Severe"
)
```
```{r analysis-glm-baseline-summary, echo=FALSE}
println("Summary of baseline model: "); 
print(liver.baseline$summary)
```

```{r analysis-glm-baseline-error, echo=FALSE}
println("Baseline classification table: ")
print(unlist(list(
  "Baseline model error rate:" = liver.baseline$error
)))
print(liver.baseline$table)
```
```{r analysis-glm-optimal-error, fig.width=10, fig.size=8, echo=FALSE}
print(liver.baseline$plot)
println("Misclassification table using optimal threshold: ")
summary(liver.baseline$optimal_table)
print(liver.baseline$optimal_table)
```

<!-- PAGE BREAK -->
\newpage

## Cross-Validation Analysis {.tabset}

### 10-fold CV

```{r analysis-glm-10cv, fig.width=10, fig.size=8, echo=FALSE}
liver.k10CV <- analysis.cv(
  liver, truth,
  params = list(
    formula = severity ~ .,
    family = "binomial"
  ),
  k = 10, rounds = 100, m = 100
)
summary(liver.k10CV)
```

```{r analysis-glm-10cv-plot, fig.width=10, fig.size=8, echo=FALSE}
print(liver.k10CV$plot)
```

```{r analysis-glm-10cv-summary, fig.width=10, fig.size=8, echo=FALSE}
printf("Optimal threshold: %s", liver.k10CV$optimal_threshold)
print(liver.k10CV$confusion_mat_summary)
```



<!-- PAGE BREAK -->
\newpage

### 3-fold CV

```{r analysis-glm-3cv, fig.width=10, fig.size=8, echo=FALSE}
liver.k3CV <- analysis.cv(
  liver, truth,
  params = list(
    formula = severity ~ .,
    family = "binomial"
  ),
  k = 3, rounds = 100, m = 100
)
summary(liver.k3CV)
```
```{r analysis-glm-3cv-plot, fig.width=10, fig.size=8, echo=FALSE}
print(liver.k3CV$plot)
```

```{r analysis-glm-3cv-summary, fig.width=10, fig.size=8, echo=FALSE}
printf("Optimal threshold: %s", liver.k3CV$optimal_threshold)
print(liver.k3CV$confusion_mat_summary)
```

<!-- PAGE BREAK -->
\newpage

## Session Information

*This document was generated from an [R Markdown](http://rmarkdown.rstudio.com) Notebook (See the `vignettes/HW3_report.Rmd` in the project's sub-directory). The setup chunk for this document sets the root directory to the project root directory using the `here` package; all file paths are relative to the project root.*

```{r session-info, echo=FALSE, message=FALSE}
sessionInfo()
```
